Quantifying the rock-paper-scissors effect

### 2021-01-20 inst/rock_paper_scissors_monster.Rmd
generalized RPS

Here I quantify the strength of a circular rock-paper-scissors effect
using the `hyper2` package, improving on previous results which merely
established its existence.

@@

We can call this the TAKM model, and find the evaluate:

@@

>>>>

Here I use the `hyper2` package (Hankin 2017) to quantify the strength
of a circular rock-paper-scissors effect using the `hyper2` package,
improving on previous results (West and Hankin 2008) which merely
established its existence.  I use two datasets, a three-player
scoreline and a synthetic dataset referring to wear on a seven-tooth
gear wheel.

@@

(in the above, a win by a player helped by the RPS effect has `M`
added to his strength).  We can call this the TAKM model, and find the
evaluate:

@@

# Generalized RPS

West and Hankin (2008) consider a synthetic dataset referring to seven
teeth of a gear wheel.  Consecutive pairs are repeatedly compared and
the most worn tooth is nominated by independent judges.  Here a
dataset of the same form will be used, but the entries have been
changed for pedagogical reasons:

```{r,echo=FALSE}
t1 <- "t1"
t2 <- "t2"
t3 <- "t3"
t4 <- "t4"
t5 <- "t5"
t6 <- "t6"
t7 <- "t7"
M <- "M"
```

```{r,label=setuphw}
hw <- matrix(c(
    16,05,NA,NA,NA,NA,NA,
    NA,06,09,NA,NA,NA,NA,
    NA,NA,05,11,NA,NA,NA,
    NA,NA,NA,06,12,NA,NA,
    NA,NA,NA,NA,05,09,NA,
    NA,NA,NA,NA,NA,04,07,
    13,NA,NA,NA,NA,NA,05),
    7,7,byrow=TRUE)
colnames(hw) <- paste("t",1:7,sep="")
hw
```

Thus, considering the top row, we see that in a total of $16+5=21$
independent comparisons between `t1` and `t2`, tooth `t1` was judged
to be the more worn one 16 times, and `t2` was judged to be the more
worn one $5$ times (NB: in the above, I have changed the dataset in
order to make `t1` stronger than the other players while maintaining
the RPS phenomenon).  We may convert this dataset to a generalized
`hyper2` object.  First a standard Bradley-Terry:

```{r,label=HWnomonster}
HW <- hyper2()
rs <- -rowSums(hw,na.rm=TRUE)
cs <- colSums(hw,na.rm=TRUE)
HW[t1] <- cs[1]
HW[t2] <- cs[2]
HW[t3] <- cs[3]
HW[t4] <- cs[4]
HW[t5] <- cs[5]
HW[t6] <- cs[6]
HW[t7] <- cs[7]
HW[c(t1,t2)] <- rs[1]
HW[c(t2,t3)] <- rs[2]
HW[c(t3,t4)] <- rs[3]
HW[c(t4,t5)] <- rs[4]
HW[c(t5,t6)] <- rs[5]
HW[c(t6,t7)] <- rs[6]
HW[c(t7,t1)] <- rs[7]
HW
```

(likelihood function `HW` has no monster).  We can then analyse this:

```{r,label=testHW}
maxp(HW)
equalp.test(HW)
```

Thus according to this model we may reject the hypothesis of equal
strength.  We now incorporate the monster:

```{r,label=NWwithmonster}
HWM <- hyper2()
tt <- c("t1","t2","t3","t4","t5","t6","t7")
for(i in 1:6){
  HWM[c(tt[i]          )] <-  hw[i,i  ]
  HWM[c(      tt[i+1],M)] <-  hw[i,i+1]
  HWM[c(tt[i],tt[i+1],M)] <- rs[i]
}
HWM[c(tt[7]        )] <-  hw[7,7]
HWM[c(      tt[1],M)] <-  hw[7,1]
HWM[c(tt[7],tt[1],M)] <- rs[7]
HWM
```

Then the same analysis applies:

```{r,label=maxpHWM}
maxp(HWM)
```

and we can test the hypothesis that $p_M=0$:

```{r}
specificp.gt.test(HWM,"M",0)
```

suggesting that we may reject the hypothesis of zero monster strength
with a p-value of about 4.7\%, even after accounting for the different
strengths of the teeth.

## References

* L. J. West and R. K. S. Hankin 2008.  "Exact tests for two-way
  contingency tables with structural zeros".  _Journal of Statistical
  Software_ 2008, volume 28 issue 11.
* R. K. S. Hankin 2010.  "A generalization of the Dirichlet
  distribution".  _Journal of Statistical Software_ 2010, volume 33
  issue 11.
* R. K. S. Hankin 2017.  "Partial rank data with the `hyper2` package:
  likelihood functions for generalized Bradley-Terry models".  _The R
  Journal_, volume 9, number 2.
